Dynamic Walking and Whole-Body Motion Planning for Humanoid Robots: an Integrated Approach

### 2012-12-07 paper.tex
Integrate comments from Florent.

There are two main ways of using motion planners to generate dynamically balanced robotic motions.
The more general one is to plan in a robot dynamic space, see for example \cite{shkolnik2011bounding}. 
By taking into account both robot
configuration and velocity, motions that satisfy dynamic balance constraints can be generated
at a planning phase. When planning motion for humanoid robots, this is a particularly costly
approach, as the size of the space to explore is augmented with the robot velocity and footstep
positions. The other way is to first plan  a geometric path that can be approximated by a
dynamic trajectory in a second step \cite{yoshida-humanoids05}. The approach we present in this paper 
falls into the second 
category. Some feasible dynamic motions are inherently impossible to compute with this kind
of approach. For example, jumping motions cannot be generated by a purely geometric planner.

\textbf{In this paper, we present a planning algorithm that considers
  exact models of a humanoid robot and its environment. It is used to
  solve navigation and manipulation problems. Our planner is a
  two-step algorithm: a first collision-free path is computed in the
  space of quasi-statically balanced configurations, then this first
  path is approximated by a sequence of dynamic walking trajectories,
  i.e. trajectories satisfying the Newton-Euler equations. The proof
  of correctness of the algorithm is based on the concept of
  small-space controllability. This property allows, under some
  assumptions, to approximate any \textit{non necessarily admissible}
  path, by a sequence of admissible trajectories. In our context we
  prove that dynamic walking makes humanoid robots small-space
  controllable. Our planner is designed for perfectly modeled indoor
  environments, where the floor is horizontal and flat. Also, because
  we do not explicitly compute footprint positions at the planning
  stage, the presented planner is unable to plan motions in which the
  robot steps over obstacles. These limitations are discussed in the
  paper.}

@@

This work is based on several fields of humanoid robotics
research: prioritized inverse kinematics, randomized  whole-body  motion
planning and walk pattern generators based on the Zero-Moment
Point (ZMP) formalism. This section summarizes the literature related to each of 
these fields.

@@

Another  field of  humanoid  robotics research  is  the generation  of
dynamically balanced walk patterns. Since  the introduction of  the ZMP
formalism  \cite{vukobratovic1969contribution},  several  methods  have  been  proposed  to
generate  walking  motions efficiently.   One  way  to  deal with  the
complexity  of a  humanoid  robot  kinematic tree  is  to use  the
so-called "cart-table" simplified model \cite{kajita2003biped}. Based on this
model,  planning a trajectory  for the  ZMP is  reduced to  planning a
trajectory  for  the Center  of  Mass (CoM)  of  the  robot.  Given  a
trajectory  of  the CoM  and  footprint  positions, inverse  kinematics
solvers can animate  the whole set of DoFs of the
robot to generate a dynamically balanced walk trajectory.

@@

Some planning methods for free-climbing robots \cite{bretl2006motion}
can be seen as a general way 
to consider quasi-static multi-step planning. They are not however directly applicable to
humanoid dynamically balanced locomotion.
Other recent contributions to the field of locomotion planning  include algorithms 
considering the dynamics at the planning phase \cite{shkolnik2011bounding}. This leads to 
a growth of algorithmic complexity, particularly costly for high-dimensional
systems such as humanoid robots. To the authors knowledge such techniques have not been
used on humanoid robotic platforms so far.

@@

Another contribution of this paper is the formal proof that ynamic
walking makes humanoid robots small-space controllable, whith the
direct implication that this first path can always be approximated by
a dynamically balanced, collision-free walking trajectory. We have
implemented this well-grounded method, and the results have been
validated on the HRP-2 robot.

@@

The problem solved here differs from classic approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the
  set of collision-free configurations satisfying a given set of
  inverse kinematics balance constraints;
\item the goal manifold $\goalmanifold$ is also defined implicitly,
  by additional inverse kinematics constraints.
\end{enumerate}
\textbf{During global planning, we will consider several types of
equality constraints for various reasons:}
\begin{itemize}
\item Static balance: the CoM of the robot stays above
  the support polygon center, the two feet have fixed position and
  orientation.
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as a specific robot hand pose,
  or a gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following section.
\end{itemize}

@@

This section presents a \textit{constructive} proof that any such statically balanced, 
collision-free path for a legged robot sliding
on the ground can be approximated by a  dynamically balanced, collision-free walk trajectory.
The proof is based on ideas from control theory, in particular small-space 
controllability. It also uses the fact that balance criteria for dynamic walking are different
from the ones for static balance. An important point is that our result and method do not hold
for a biped robot that walks in a quasi-static way.

@@

The algorithm that animates a statically balanced path into a
dynamically balanced walk trajectory has been inspired by the previous
small-space controllability proof. \textbf{Given a statically balanced
  path $p$ verifying the cart-table model approximation constraints,
  we start by placing footprints corresponding to the nominal walk
  pattern of the robot. Given the footprints, we compute a ZMP
  trajectory, derive foot trajectories, and a preview controller
  outputs a corresponding CoM trajectory. Classic numerical
  Jacobian-based prioritized inverse kinematics methods prove to be
  very useful to generate a dynamic walking trajectory while trying to
  accomplish secondary tasks, such as following a reference
  configuration trajectory. We use the same framework which was
  developed in \cite{yoshida2006tds}.}

The stack of tasks applied to the robot to generate a dynamic walking
motion is -- in decreasing priority order:

@@

\STATE $StackOfTasks$.initialize()

\STATE $StackOfTasks$.addFootprintTask($Footprints$)

\STATE $StackOfTasks$.addWaistTask()

\STATE $StackOfTasks$.addConfigurationTask($p$)

\STATE $DynamicTrajectory \leftarrow
\text{ComputeWalkTrajectory}(StackOfTasks)$

@@

@@

\textbf{The presented experiment setup assumes perfect knowledge of
  the environment. This can be guaranteed during experiments by using
  calibrated objects and motion capture systems. This indeed allows us
  to focus on complex motion planning problems. The perception
  problem, interesting as it is, is thus completely decoupled from the
  planning problem in our presentation.}

\textbf{Our constrained motion planner and the small-space
  controllability property can luckily still be used with different
  environment representations that might be more suited for perception
  systems. Extensions \cite{Nakhei4755945,DanLauLam2012} are currently
  under development to relax the a priori environment knowledge
  constraint and achieve reactive planning in unknown environments.}

@@

The presented setup also assumes perfect execution of the plan. It can
be critical here, since non-nominal stepping may cause the robot to
drift away from the planned trajectory. Future experiments will
include trajectory following during plan execution.

>>>>

There are two main ways of using motion planners to generate dynamically balanced robotic motions.
The more general one is to plan in a robot dynamic space, see for example \cite{shkolnik2011bounding}. 
By taking into account both robot
configuration and velocity, motions that satisfy dynamic balance constraints can be generated
at a planning phase. When planning motion for humanoid robots, this is a particularly costly
approach, as the size of the space to explore is augmented with the robot velocity and footprint
positions. The other way is to first plan  a geometric path that can be approximated by a
dynamic trajectory in a second step \cite{yoshida-humanoids05}. The approach we present in this paper 
falls into the second 
category. Some feasible dynamic motions are inherently impossible to compute with this kind
of approach. For example, jumping motions cannot be generated by a purely geometric planner.

\textbf{In this paper, we present a planning algorithm that considers
  exact models of a humanoid robot and its environment. It is used to
  solve navigation and manipulation problems. Our planner is a
  two-step algorithm: a first collision-free path is computed in the
  space of quasi-statically balanced configurations, then this first
  path is approximated by a sequence of dynamic walking
  trajectories. The proof of correctness of the algorithm is based on
  the concept of small-space controllability. This property allows,
  under some assumptions, to approximate any \textit{non necessarily
    admissible} path, by a sequence of admissible trajectories. In our
  context we prove that dynamic walking makes humanoid robots
  small-space controllable. Our planner is designed for perfectly
  modeled indoor environments, where the floor is horizontal and
  flat. Also, because we do not explicitly compute footprint positions
  at the planning stage, our planner is unable to plan motions in
  which the robot steps over obstacles. These limitations are
  discussed in the paper.}

@@

This work is based on several fields of humanoid robotics research:
prioritized inverse kinematics, randomized whole-body motion planning
and walk pattern generators based on the Zero-Moment Point (ZMP). This
section summarizes the literature related to each of these fields.

@@

Another field of humanoid robotics research is the generation of
dynamically balanced walk patterns. Since the introduction of the ZMP
\cite{vukobratovic1969contribution}, several methods have been
proposed to generate walking motions efficiently.  One way to deal
with the complexity of a humanoid robot kinematic tree is to use the
so-called "cart-table" simplified model \cite{kajita2003biped}. Based
on this model, planning a trajectory for the ZMP is reduced to
planning a trajectory for the Center of Mass (CoM) of the robot.
Given a trajectory of the CoM and footprint positions, inverse
kinematics solvers can animate the whole set of DoFs of the robot to
generate a dynamically balanced walk trajectory.

@@

Some planning methods for free-climbing robots \cite{bretl2006motion}
can be seen as a general way to consider quasi-static multi-step
planning. They are not however directly applicable to humanoid
dynamically balanced locomotion.  Other recent contributions to the
field of locomotion planning include algorithms considering the
dynamics at the planning phase \cite{shkolnik2011bounding}. This leads
to a growth of algorithmic complexity, particularly costly for
high-dimensional systems such as humanoid robots. To the authors'
knowledge such techniques have not been used on humanoid robotic
platforms so far.

@@

Another contribution of this paper is the formal proof that dynamic
walking makes humanoid robots small-space controllable, with the
direct implication that this first path can always be approximated by
a dynamically balanced, collision-free walking trajectory. We have
implemented this well-grounded method, and the results have been
validated on the HRP-2 robot.

@@

The problem solved here differs from classic approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the
  set of collision-free configurations satisfying a given set of
  inverse kinematics balance constraints;
\item the goal manifold $\goalmanifold$ is also defined implicitly,
  by additional inverse kinematics constraints.
\end{enumerate}
During global planning, we will consider several types of constraints
for various reasons:
\begin{itemize}
\item \textbf{Static balance: the CoM of the robot stays above the support
  polygon center, the two feet are horizontal on the ground.}
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as a specific robot hand pose,
  or a gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following section.
\end{itemize}

@@

This section presents a \textit{constructive} proof that any such statically balanced, 
collision-free path for a legged robot sliding
on the ground can be approximated by a  dynamically balanced, collision-free walk trajectory.
The proof is based on ideas from control theory, in particular small-space 
controllability. It also uses the fact that balance criteria for dynamic walking are different
from the ones for static balance.

@@

The algorithm that animates a statically balanced path into a
dynamically balanced walk trajectory has been inspired by the previous
small-space controllability proof. \textbf{Given a statically balanced
  path $p$ verifying the cart-table model approximation constraints,
  we start by placing footprints corresponding to the nominal walk
  pattern of the robot. Given the footprints, we compute a ZMP
  trajectory, derive foot trajectories, and a preview controller
  returns the corresponding CoM trajectory. Classic numerical
  Jacobian-based prioritized inverse kinematics methods prove to be
  very useful to generate a dynamic walking trajectory while trying to
  accomplish secondary tasks, such as following a reference
  configuration trajectory. We use the framework called Generalized
  inverse kinematics (Gik) and developed in \cite{yoshida2006tds}.}

\textbf{The hierarchy of tasks (referred to as \textit{GikTasks} in
  Algorithm ~\ref{alg:walk}) applied to the robot to generate a
  dynamic walking motion is -- in decreasing priority order:}

@@

\STATE $GikTasks$.addFootprintTask($Footprints$)

\STATE $GikTasks$.addWaistTask()

\STATE $GikTasks$.addConfigurationTask($p$)

\STATE $DynamicTrajectory \leftarrow
\text{ComputeWalkTrajectory}(GikTasks)$

@@

To get more natural motions in the experiments, we require the foot
positions to be fixed with respect to each other, and the CoM to be
projected in the center of the support polygon during the sliding path
planning stage.

@@

\textbf{The experimental setup assumes perfect knowledge of the
  environment. This can be guaranteed during experiments by using
  calibrated objects and motion capture systems. This indeed allows us
  to focus on complex motion planning problems. The perception
  problem, interesting as it is, is thus completely decoupled from the
  planning problem in our presentation. From previous experiences
  however, we are confident that our algorithm will work as well in
  environments modeled by vision \cite{Nakhei4755945,DanLauLam2012}.}

@@

The setup also assumes perfect execution of the plan. It can be
critical here, since non-nominal stepping may cause the robot to drift
away from the planned trajectory. Future experiments will include
trajectory following during plan execution.
